Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-07 survey/survey.tex
sec 2

For convenience, we write $[2^n]$ for the finite set representing
$n$-bit words with the canonical ordering for binary numbers. Thus
when $n=3$, the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which
correspond to the 3-bit words $\{000,001,010,011,100,101,110,111\}$.
Although this language appears weak, it is universal for boolean
combinational circuits. To see this, we express, boolean negation
\emph{not} as a permutation $[2^1] \rightarrow [2^1]$, controlled not
(also known as \emph{cnot}) as a permutation
$[2^2] \rightarrow [2^2]$, and controlled controlled-not (also known
as \emph{toffoli}) as a permutation $[2^3] \rightarrow [2^3]$:
\[\begin{array}{rcl}
\mathit{not} &=& \mathit{swap}~0~1 \\
\mathit{cnot} &=& \mathit{swap}~2~3 \\
\mathit{toffoli} &=& \mathit{swap}~6~7
\end{array}\]
For example, the \emph{toffoli} gate negates the third bit (the target
bit) if both the first two bits (the control bits) are 1, i.e., it
swaps $110$ and $111$. Universality follows since any combinational
circuit can be realized using compositions of \emph{toffoli} gates.

There is however a subtle issue: programming in such an unstructured
language is \emph{not} compositional in the sense that using
$\mathit{not}$ in a larger circuit forces us to change its
implementation. To illustrate this point, consider the reversible full
adder below designed by Desoete et al.~\cite{117414}:

@@

In the figure (copied from a more general paper that includes
alternative designs~\cite{Rentergem2005OptimalDO}), the full adder
takes 4 inputs: the two bits to add $A$ and $B$, an incoming carry bit
$C_i$, and a heap input initialized to 0 to maintain
reversibility. There are four outputs: the first two are identical to
the incoming bits $A$ and $B$ and are considered ``garbage.'' The
third bit $S$ is the sum and the last bit $C_o$ is the outgoing carry
bit. In the notation used to describe the circuit, the $x$ denotes
boolean negation, the dots are control bits. In our reversible
language, we can express this circuit as the following permutation of
type $[2^4] \rightarrow [2^4]$:

\[\begin{array}{ll}
\mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 & \mbox{toffoli} \\
\mathit{swap}~8~12 \fatsemi \mathit{swap}~9~14 \fatsemi 
    \mathit{swap}~10~13 \fatsemi \mathit{swap}~11~15 & \mbox{cnot and swap} \\
\mathit{swap}~6~7 \fatsemi \mathit{swap}~14~15 & \mbox{toffoli} \\
\mathit{swap}~4~6 \fatsemi \mathit{swap}~5~7 \fatsemi
    \mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 & \mbox{cnot}
\end{array}\]

Note how the implementation of $\mathit{cnot}$ as a permutation
$[2^2] \rightarrow [2^2]$ cannot be directly reused in a circuit
$[2^4] \rightarrow [2^4]$. Indeed, in programming practice we are
interested in structured data and compositional abstractions which
will be the subject on next section. What we do learn however from
this short investigation using untyped flat sets is what the
``\emph{purely operational}'' view of the theory would be. In
particular, it tells us that permutations are an inescapable part of
the fabric of reversible computing.  However as permutations are
untyped, and act on the canonicalized version of $n$-element sets
(i.e. those sets where all the structure has been forgotten), these
are a rather pale shadow of the rich tapestry of
information-perserving transformations of structured data which we
investigate next.

>>>>

For convenience, we write $[2^n]$ for the finite set representing
$n$-bit words with the canonical ordering for binary numbers. Thus
when $n=3$, the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which
correspond to the 3-bit words $\{000,001,010,011,100,101,110,111\}$.
Although this language appears weak, it is universal for reversible
boolean combinational circuits $[2^i] \rightarrow [2^i]$ with $i$
input/output wires.

We illustrate the expressiveness of the language, we develop a few
small examples. We start by writing boolean negation ``not'' as a
permutation $[2^1] \rightarrow [2^1]$, the controlled-not gate (also
known as ``cnot'') as a permutation $[2^2] \rightarrow [2^2]$, and the
controlled-controlled-not gate (also known as ``toffoli'') as a
permutation $[2^3] \rightarrow [2^3]$:
\[\begin{array}{rcl}
\mbox{not} &=& \mathit{swap}~0~1 \\
\mbox{cnot} &=& \mathit{swap}~2~3 \\
\mbox{toffoli} &=& \mathit{swap}~6~7
\end{array}\]
The ``cnot'' gate operates on two bits and negates the second (the
target bit) if the first one (the control bit) is 1, i.e., it swaps
$10$ and $11$; the ``toffoli'' gate negates the third bit (the target
bit) if both the first two bits (the control bits) are 1, i.e., it
swaps $110$ and $111$.

There is however a subtle issue: programming in such an unstructured
language is \emph{not} compositional in the sense that using the
``not'' gate in a larger circuit forces us to change its
implementation. Indeed if we had two bits and wanted to use ``not'' to
negate the first bit, we would write the permutation of type
$[2^2] \rightarrow [2^2]$ that permutes $00$ with $10$ \emph{and}
permutes $01$ with $11$, i.e, the permutation
$\mathit{swap}~0~2 \fatsemi \mathit{swap}~1~3$. To further illustrate
how inconvenient this point is, consider the reversible full adder
below designed by Desoete et al.~\cite{117414}:

@@

In the figure (copied from a more general paper that includes
alternative designs~\cite{Rentergem2005OptimalDO}), the full adder
takes 4 inputs: the two bits to add $A$ and $B$, an incoming carry bit
$C_i$, and a heap input initialized to 0 to maintain
reversibility. There are four outputs: the first two are identical to
the incoming bits $A$ and $B$ and are considered ``garbage.'' The
third bit $S$ is the sum and the last bit $C_o$ is the outgoing carry
bit. In the notation used to describe the circuit, the $\times$ denotes
boolean negation, the dots are control bits. In our reversible
language, we can express this circuit as the following permutation of
type $[2^4] \rightarrow [2^4]$:

\[\begin{array}{ll}
\mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 ~\fatsemi & \mbox{toffoli} \\
\mathit{swap}~8~12 \fatsemi \mathit{swap}~9~14 \fatsemi 
    \mathit{swap}~10~13 \fatsemi \mathit{swap}~11~15 ~\fatsemi & \mbox{cnot and swap} \\
\mathit{swap}~6~7 \fatsemi \mathit{swap}~14~15 ~\fatsemi & \mbox{toffoli} \\
\mathit{swap}~4~6 \fatsemi \mathit{swap}~5~7 \fatsemi
    \mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 & \mbox{cnot}
\end{array}\]

Note how the implementation of $\mathit{cnot}$ as a permutation
$[2^2] \rightarrow [2^2]$ cannot be directly reused in the larger
circuit $[2^4] \rightarrow [2^4]$.

For such reasons, in programming practice we are interested in
structured data and compositional abstractions which will be the
subject of the next section. What we do learn however from this short
investigation using untyped flat sets is what the ``\emph{purely
  operational}'' view of the theory would be. In particular, it tells
us that permutations are an inescapable part of the fabric of
reversible computing.  However as permutations are untyped, and act on
the canonicalized version of $n$-element sets (i.e. those sets where
all the structure has been forgotten), these are a rather pale shadow
of the rich tapestry of information-perserving transformations of
structured data which we investigate next.
